\subsection{Software}
We developed a software package with a user interface that implements the above LP problem and allows configuration of clinicians at [ref \ref{???}], to be used by the ID division at St. Michael's Hospital. The software was used to generate the results in the following sections, using real data as well as simulated data as input.

\subsection{Infectious Diseases Division}
We used clinician time-off requests and minimum/maximum requirements from 2015-2018 as input data for the LP problem. Tables \ref{tbl:2018-schedule-comparison}, \ref{??}, \ref{??} present the optimal schedule generated using the software as well as the manually created schedule, color-coded to distinguish between the different clinicians assigned.

\include{tbl/2018_schedule_comparison}

In order to evaluate the generated schedule and compare it with the manually created schedule, we outline the adherence of each schedule to the constraints presented in section \ref{???}. As we can see, the generated schedule was able to satisfy all mandatory constraints, however for the years of [???] it could not find an optimal schedule while ensuring that the block assignments are spread out ([???] constraint). On the other hand, we can see that the manually created schedule was not able to satisfy all mandatory constraints. In particular, we see that it contains consecutive block assignments for the years of [???], [...]. Evaluating the objectives, we see that [...\textit{request conflicts}]. Moreover, the manual schedule does not attempt to align weekend assignments with block assignments, unlike the optimal schedule found by the software.

\subsection{Simulations}